Scaling up a GF grammar for free parsing

### 2011-12-12 doc/report/report.tex
better version of ch.1&2

\tableofcontents
\newpage
\chapter{Introduction}
%>> short intro to GF
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that give a deep analyse for
Swedish. The fact that the existing rule-based Swedish
parsers are not freely available does not only restrict their
usage but also their possiblities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.\\
To build a parser or a grammar from scratch is a heavy work ~~ and we
therefore combine already existing resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

\section{Outline}
This thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} and a deeper
presentation of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction Swedish and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

\chapter{Background}  
\label{sec:background}
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work \ref{??} the start on extending the Swedish GF grammar was
made and a tool for lexical aquistion were developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in this section.

\section{Swedish}
\label{sec:swedish}
>> move, have GF first?\\
Swedish \cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
an official language in Finland and altogether spoken by approximately 10 ~~
9?Check SAG million people.
Swedish syntax is often similar to English, but have some characteristics ~~:

\subsection*{Post-nominal articles} 
Unlike most of the world's languages Swedish express not only the number but also
the definitness of a noun by suffixes. The endings are decided by the gender of the noun,
which may be \textit{neuter} or \textit{common}.\\
%(or \textit{utrum})

\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{common}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

The definiteness can be accentuated by using the definite articles.\\

\begin{tabular}{llll}
den katten & det huset & de katterna & de husen \\
\emph{the cat} & \emph{the house} & \emph{the cats} & \emph{the houses} \\
\end{tabular}\\

The definite articles are required when the noun is modified by adjectival phrase, sentence
\ref{sent:trottkatt}b is hence not grammatically correct.
%\enumsentence{\shortexnt{4}
%{Den& trötta& katten & sov}
%{the& tired& cat+\textbf{the} & slept}}
\enumsentence{\begin{tabular}{llllll}
a.& Katten sov. & b.& *Trötta katten sov &c. &Den trötta katten sov\\
&The cat slept. & &~~The tired cat slept& &The tired cat slept\\
\end{tabular}}\label{sent:trottkatt}

%can be superdoubled
%example: den där boken

\subsection*{Verb second}
Swedish is a verb-second language \cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted \cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, but vary in how the content is presented.
%~~ although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being
%emphasized.
\enumsentence{
\shortex{4}
{Du & ser &  inte &mig.}
%empha glossning!! ha samma överallt, ' eller ''
{you & see & not & me}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{me & see & you & not}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{not&  see& you& me}} \label{ex:swedish-avso}

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{saw  & you&  me& not}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%>> and stricter
\enumsentence{Main sentences:\\
{\shortex{8}{Jag & såg & Anna & men & hon & såg & {\bf inte} & mig }
{ I & saw & Anna & but & she & saw & {\bf not} & me}
{`I saw Anna but she didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & förstod & att& Anna& \textbf{inte}& såg& mig.}
            {I &  understood & that& Anna& \textbf{not}& saw& me}
            {`I understood that Anna didn't see me'}}

\subsection*{Passive voice}
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
%>> better with animate subjects
\enumsentence{
\shortex{5}
{Tjuven &blev& tagen& av & polisen}
{the thief & was & taken & by & the police }
{`The thief was arrested by the police'}}
\label{gfPass:peri-pass}
and the \textbf{s-passive} which is formed by adding an \emph{s} to the verb: \\
\enumsentence{Passive \hspace{45mm} Active\\
\shortexm{12}
{Tjuven & togs&  av & polisen}
{the thief&  took+\textbf{s}&  by&the police}
{`The thief was arrested by the police'}
{&&Polisen & tog & tjuven}
{&&the police & took & the thief}
{\hspace{-9mm}`The police arrested the thief'}}
\label{gfPass:s-pass}
The s-passive is the most commonly used one for both written
and Swedish \cite[??]{SAG} and dominates especially when the subject is inanimate.
%cf.
%\enumsentence{
%\shortex{5}{Ett & lejon & jagade & honom.}
%        {A &lion &hunted &him.}
%        {`A lion was hunting him.'}}

\subsection*{Impersonal constructions}
Constructions with `det är/var' (\emph{`it is/was'}) is very common in Swedish \cite[\textsection 309d]{H&H}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{it &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject in presentational constructions where
the real subject is put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ it & stands &a & moose &on &the field}
{`There is a moose in the field.'}}

@@

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The ~~words are equipped ~~ with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word has different meanings, there will be separate entries for each of
them.  The word ``växa" (``grow") for example, has two meanings: \emph{getting
bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morphological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that do not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and the formalism is 
%equivalent to Parallel Multiple Context-Free Grammars (Seki \& al. 1991), an instance of
%generalized context free grammar (GCFG) it is hence
stronger than mildly contex-free grammars and hence has the same expressiveness
as Tree Adjoining Grammars (Joshi,1975) and Head Grammars (Pollard,1984).

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. The logic and reasoning 
are inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented by the Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. On of the biggest
achivements is a library covering the 
basic morphological and syntactical structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.3\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}

>>>>

\tableofcontents
\newpage
\chapter{Introduction}
% more intro to gf? Good starting sentence.
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep analyse for
Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possiblities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
made use of in other projects.\\
To build a parser or a grammar from scratch would be not only time conusming but
also a great theoretical and techniqual challenge. In order to not reinvent the wheel
we develop our parser by combining well-tested resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction Swedish, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

\chapter{Background}  
\label{sec:background}
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work\footnote{??}, the start on extending the Swedish GF grammar was
made and a tool for lexical aquistion was developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one defines a parser, and we develop it by getting
examples, ideas and test material from the treebank Talbanken.
The project is hence heavily depending on three resources, which will be described
in this section.

\section{Grammatical Framework}
\label{sec:gf}
The main component of the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and represents a formalism
%equivalent to Parallel Multiple Context-Free Grammars (Seki \& al. 1991), an instance of
%generalized context free grammar (GCFG) it is hence
stronger than mildly contex-free grammars. The framework has  hence the same expressiveness
as Tree Adjoining Grammars (Joshi,1975) and Head Grammars (Pollard,1984).

GF is a strongly typed functional programming language, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. It is also a logical framework,
and the built functionality for logic and reasoning 
is inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented by the Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. On of the biggest
achivements is a library covering the 
basic morphological and syntactical structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.22\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}
The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%A GF parse tree has two visualized representations, the \textit{parse tree} and
%the \textit{abstract tree}.
The \textit{parse tree} (\ref{pic:gfStree}) shows which types
that are assigned to the words and phrases. 
The \textit{abstract tree} shows which functions GF uses
to parse a sentence.
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The reader should keep in mind that the visualized parse trees is not a
complete representation. It does not reproduce all information from the abstract tree,
%If the end node of a  branch has no corresponding word, it is excluded from
%the visualized parse tree. This is the case for the 
eg. the definitnes and number of the noun \emph{`katten'}, which is shown as
\verb-DetQuant DefArt NumSg- in the abstract tree.
%and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the article \emph{``the"} is explicit and therefore shows as a quantifier in the parse tree.\\

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inhereted ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has succesfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in wich GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{molto} for online translation between up to 15 language.\\

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
An abstract grammar can be implemented by a set of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as fstructures and cstructures in Lexical Functional Grammar (Bresnan and Kaplan, 1982).

The example in figure \ref{fig:gfAbstract1} shows an abstract grammar defining three categories, % \verb|Categories|,
one for nouns, one for verbs and one for sentences. The abstract grammar also gives
the function types. In this case we have \verb|Pred|, which tells us that by
taking a noun and a verb we can form a sentence.  No information of how this is
done is given, at this stage. The grammars also defines two words, the noun
\verb|cat_N| and the verb \verb|sleep_V|.

%>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. We add a category \verb|N'| to the
abstract, which represents a noun with a fixed number,
%Note that \verb-N'- is normally used for nouns with a fixed definitness, but for 
%now we concentrate on the number only. Hence \verb|N| now means a noun in any of the number,
and we introduce two function, \verb|NSg : N -> N'| and
\verb|N'l : N -> N'| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; N' ; V ; S ;               lincat V, S, N' = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : N' -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> N' ;                    NPl n = n.s ! Pl ;
    NPl : N -> N' ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records, tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s|, \verb-Num => Str- shows that it is a table, which given a
parameter of type \verb|Num| returns a string. \verb|Num| is defined 
%as a parameter, which can
to either have value \verb|Sg| or \verb|Pl|. 
The dot (\verb-.-) is used for projection and the bang (\verb-!-) as a selection operator.
%In \verb|NPl| and \verb|NSg|,
\verb|n.s ! Sg| thus means that we use the branch for \verb-Sg- in
field \verb|s| of \verb|n|.
%, and then the selection operator \verb|!| is used to select a branch in the table.

When implementing a English version of the grammar, we encounter another
problem: the verb form depends on the number of the noun. The \verb|N'| needs to carry
information about which its number and \verb|Pred| must pass this on to the
verb. Finally, the type of \verb|V| needs a table, depending on the number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S = Str ;
         V = {s : Num => Str} ;
         N = {s : Num => Str} ;
         N' = {s : Str ; num : Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {s = n.s ! Pl ; num = Pl} ;
    NSg n = {s = n.s ! Sg ; num = Sg} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

We now have two implementations of the abstract. The resulting GF grammar is able both to
parse a string to a abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a advantageous resource library \cite{gf-resource}, covering the
fundamental morphology and syntactics of more than 20 languages.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and are valuable
%they constitute  a very useful resource 
when implementing application grammars. There is also a small test lexicon
included, containing a few hundred common words.

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easygoing manners are added
and the possibilities to use GF in different environments
increased. There are research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ramona} or generating natural
language via Montague semantics \cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar to which the result has a strong connection.

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large-scale treebank,
Talbanken \cite{talbanken} was perfect for our purpose.
%It was created in the 1970s at Lund University.
In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser \cite{malt}
and was then redistributed in a more modern version, Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
XML-formats
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
For each word there is semantical, syntactical and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

Words with multiple meanings are separated inte different entries.
%If one word has different meanings, there will be separate entries for each of them. 
The word ``växa" (``grow") for example, has two meanings: \emph{getting
bigger}
\enumsentence{Gräset växte varje dag\\
`The grass grew each day'}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
`There stood an old oak.'}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morphological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that do not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

\section{Swedish}
\label{sec:swedish}
Swedish \cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
one of the official languages in Finland and altogether spoken by approximately 9
million people.
Swedish syntax is often similar to English, but the  morphology is richer and the
word order slightly more intricate.

\subsection*{Post-nominal articles} 
Unlike most of the world's languages Swedish express not only the number but also
the definitness of a noun by suffixes. The endings are decided by the noun's gender,
%which may be 
\textit{neuter} or \textit{common}.\\
%(or \textit{utrum}) 
\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{common}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

The definiteness can be accentuated by using the definite articles.\\
\begin{tabular}{llll}
den katten & det huset & de katterna & de husen \\
\emph{the cat} & \emph{the house} & \emph{the cats} & \emph{the houses} \\
\end{tabular}\\

A definite article or a determiner is necessary when the noun is modified by adjectival
phrase; sentence \ref{sent:trottkatt}a is not grammatically correct.
Since adjectives also marks definitness, it is marked on three places
in sentence \ref{sent:trottkatt}b. %, the definiteness is 
%The definiteness in then marked by the article, on the noun and on the adjective.

\enumsentence{\begin{tabular}[t]{@{}*{10}{l@{\ }}}   %\begin{tabular}[t]{@{}*{lllllll}}
a. & *Gamla          & katten        &sov. &&b. &Den             &gamla            & katten         &sov.\\
   & $[${\sc+def}$]$ & $[${\sc+def}$]$&    &&  & $[${\sc+def}$]$ & $[${\sc+def}$]$ & $[${\sc+def}$]$&  \\
   &~~`The old       & cat           &slept.'&& &`The            & old             & cat            &slept.'\\
\end{tabular}}\label{sent:trottkatt}
However, for some determiners, ie. `min' (\emph{`my'}), the noun should be in indefinite form.
%the definitness of the noun is set by the determiner. For some,
%like `min' , the noun should be used in indefinite form.

\enumsentence{\longexnt{3}{3}
{Min &trötta& katt.} %&& Den där & trötta &katten}
{`My & tired & cat'}
{ $[${\sc+def}$]$ &$[${\sc+def}$]$& $[${\sc-def}$]$} % . && $[${\sc+def}$]$  & $[${\sc+def}$]$ &$[${\sc+def}$]$}
{}}

%a.& Katten& sov. & b
% &$[${\sc+def}$]$ & %can be superdoubled
%  &`The cat& slept.'%example: den där boken

\subsection*{Verb second}
Swedish is a verb-second language \cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted \cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, but vary in how the content is presented.
%~~ although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being
%emphasized.
\enumsentence{
\shortex{4}
{Du & ser &  inte &mig.}
%empha glossning!! ha samma överallt, ' eller ''
{\emph{you} & \emph{see} & \emph{not} &\emph{ me}}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{\emph{me} & \emph{see} & \emph{you} &\emph{ not}}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{\emph{not}&  \emph{see}& \emph{you}& \emph{me}}} \label{ex:swedish-avso}

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{\emph{saw}  & \emph{you}&  \emph{me}& \emph{not}}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly modified,
%>> and stricter
\enumsentence{Main sentences:\\
{\shortex{8}{Jag & såg & Anna & men & hon & såg & {\bf inte} & mig }
{ \emph{I} & \emph{saw} & \emph{Anna} & \emph{but} & \emph{she} & \emph{saw} & {\bf \emph{not}} &\emph{ me}}
{`I saw Anna but she didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & förstod & att& Anna& \textbf{inte}& såg& mig.}
            {\emph{I} &  \emph{understood} & \emph{that}& \emph{Anna}& \textbf{\emph{not}}& \emph{saw}& \emph{me}}
            {`I understood that Anna didn't see me'}}

\subsection*{Passive voice}
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
%>> better with animate subjects
\enumsentence{
\shortex{5}
{Tjuven &blev& tagen& av & polisen}
{\emph{the} \emph{thief} & \emph{was} & \emph{taken} & \emph{by} & \emph{the police} }
{`The thief was arrested by the police'}}
\label{gfPass:peri-pass}
and the \textbf{s-passive} which is formed by adding an \emph{s} to the verb: \\
\enumsentence{Passive \hspace{45mm} Active\\
\shortexm{12}
{Tjuven & togs&  av & polisen}
{\emph{the}\emph{ thief}&  \emph{took}+\textbf{s}&  \emph{by}&\emph{the police}}
{`The thief was arrested by the police'}
{&&Polisen & tog & tjuven}
{&&\emph{the police} & \emph{took} & \emph{the thief}}
{\hspace{-9mm}`The police arrested the thief'}}
\label{gfPass:s-pass}
The s-passive is the most commonly used one for both written
and Swedish \cite[Pass. \textsection 1]{SAG} and dominates especially when the subject is inanimate.
%cf.
%\enumsentence{
%\shortex{5}{Ett & lejon & jagade & honom.}
%        {A &lion &hunted &him.}
%        {`A lion was hunting him.'}}

\subsection*{Impersonal constructions}
Constructions with `det är/var' (\emph{`it is/was'}) is very common in Swedish \cite[\textsection 309d]{H&H}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{\emph{it} &\emph{was} & \emph{nice} &\emph{to} &\emph{hear}.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject in presentational constructions where
the real subject is put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ \emph{it} & \emph{stands} &\emph{a} & \emph{moose} &\emph{on} &\emph{the field}}
{`There is a moose in the field.'}}

@@
